Automatic Parallelization and Transparent Fault Tolerance\\ (New project paper)

### 2016-04-01 pubs/tfp16/paper.tex
minor edits

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
% mkd for preliminary paper notes
\usepackage{enumitem} 
\usepackage{color} % for parenthetical notes while paper is in progress
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Strict Pure Functional Programming and Automatic Parallelization and Transparent Fault Tolerance\\
       (New project paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{abstract}
The claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Our project is the demonstration
of a light-weight, higher-order, polymorphic, pure functional language
implementation in which we can experiment with automatic parallelization
strategies and varying degrees of default strictness.  Non-strictness
constrains the degree of potential parallelism, that is, we are not exploring
speculative evaluation.  A secondary goal is to experiment with mechanisms
for transparent fault tolerance.
\\
\emph{Draft paper notes:}
\begin{enumerate}[before=\itshape]
  \item We are reporting work in progress.  At the time of writing we have a
    complete serial implementation, a primitive proof-of-concept parallel
    implementation, and are currently developing a more realistic parallel
    runtime.  We expect to have preliminary results from our new parallel
    implementation by the time of the symposium.  The fault-tolerance mechanisms
    remain future work.

@@

  \item Hopefully in the spirit of a student-friendly symposium, for this
    draft version we have chosen a more informal, conversational style of
    presentation than necessarily appropriate for a final paper, have
    highlighted student involvement and its influence on our approach, and
    included perhaps gratuitous detail about process.
  \end{enumerate}

\keywords{strict pure functional programming, automatic parallelization, Haskell, STG}

@@

\subsection{The trend: strict but still pure functional programming in HPC}
Interest in functional programming for HPC in general is not new, as evidenced
by the Workshop on Functional High Performance Computing, held in conjuction
with the International Conference on Functional Programming, in its fifth
consecutive year.  A more precise characterization of the trend is that there
is a small but growing understanding by practicing computational scientists
that \emph{pure functional semantics}, in some form, is essential to reining
in the complexity of these ever-evolving scientific codes and informing the
designs of new ones.  Such semantics are expressed in a number of ways as
highlighted following.

@@

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practioners who are developing an appreciation for pure functional programming
in Haskell.  Still, they see their needs as different than other Haskell
users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
primary data structure, and non-strictness or laziness is more of a
nuisance--even a serious hindrance--than a convenience, much less ever
algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
Johnsson's general attribute grammars~\cite{Johnsson87}, or other instances
of \emph{tying the knot.}~\cite{tying-the-knot}.

A common complaint among these aspiring high-performance functional
programmers is the need to get strictness just right: to remove space leaks,
achieve acceptable performance, get communication (e.g.\ MPI) and IO to work
as they expect, and so on.  More specifically, they expect to be able to
reason about space and time usage, and order of execution, using their
existing mental models. While they are happy to write code that is strongly
reminiscent of the numerical examples in Hughes' classic piece~\cite{Hughes89}
(whether they've read it or not), they would also be happy to write in a
slightly modified style that does not require non-strictness---after all,
that's how they've been thinking their entire careers.

Much anecdotal evidence suggests that many others have similar difficulties
with semantics that are lazy (or non-strict) by default, with the most telling
recent concrete evidence being the push for, and realization of, the
\texttt{Strict} and \texttt{StrictData} language extensions newly available in
GHC 8.0.  Quoting the documentation:
\begin{quote}
  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
  with bang patterns, making it harder to read. The reason is that laziness
  isn't the right default in this particular code, but the programmer has no
  way to say that except by repeatedly adding bang
  patterns~\cite{strict-strictdata}.
\end{quote}
It is a fact that \emph{HPC practioners are by and large not computer
  scientists}, they are domain scientists (physicists, chemists, etc.) or
computational scientists specialized in numerical or numerically-oriented
algorithmic methods (e.g., Lagrangian, Eulerian, adaptive mesh refinement).
The concept of a space leak, for example, caused by other than a missing
explicit deallocation statement (C \emph{free}, etc.) is difficult to
understand and not something they particularly want to try to understand.
Hence, we eschew explicit strictness annotations for varying uniform degrees
of strictness.

More generally, \emph{task-based} parallelism is becoming increasingly popular
for HPC, often in conjunction with a higher-level (typically inter-node) bulk
synchronous model typically implemented with the Message Passing Interface
(MPI)~\cite{MPI}.  In the imperative world, example systems include the
aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
and the HPX C++11/14 extension~\cite{HPX}.  While the concept of task is
somewhat vague, the general idea is a unit of computation with no or
minimal/constrained side effects.  Function call in a pure functional language
could be regarded as the essence of the concept of task.

In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack.  Current practice in HPC is checkpoint/restart, wherein at
various point in time sufficient program state is dumped to backing store such
that the program, in case of a fault, can be (manually) restarted from the
latest stored state.  Most implementations are explicit, though there has been
some success with transparent (invisible to the programmer) systems.  Wiring
checkpoint/restart into a large-scale application is not only burdensome and
error-prone, it can severely warp the engineering of a large-scale application
because it must be designed such that essential state can be readily captured.
And now, it is widely believed that the scalability of checkpoint/restart may
be reaching its end.

In the imperative world side-effect-free tasks could be safely restarted after
failure, needing only their well-defined inputs and not arbitrary access to
mutable global state.  The Legion system is being augmented with this
capability, and in such a way that it will be largely transparent to the
programmer.  Ericsson's quasi-functional language Erlang and its runtime
system was designed for fault tolerance.~\cite{Cesarini:2009}.  It has long
been recognized that the pure parts of a functional program (excluding, e.g.,
the I/O monad in Haskell) could be safely restarted.  The recent work of
Steward impressively demonstrates this on a large-scale distributed-memory
system~\cite{Stewart:2013}.

We claim, then, that there is a trend, in HPC at least, towards
\emph{strict-by-default pure functional programming}, whether as imposed by
programmer discipline, a parallel runtime system, or the language itself
(e.g.\ strict Haskell).  Second, we claim that there is a trend in
high-performance computing, and by transitivity in functional programming,
towards transparent fault tolerance.

@@

\section{Project and Goals}
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with degrees of language
strictness ranging from lazy to fully strict, and secondarily, with mechanisms
for transparent fault tolerance.  Light-weight has several meanings or
implications.  First and foremost it means that \emph{we are not attempting to
  compete with GHC in any respect}---succumbing to the temptation of
recreating GHC language, type system, and runtime features, or significant
code optimization, would simply divert us from our central goals.\footnote{We
  estimate that we will have, to within an order of magnitude, one person-hour
  of total available development time for each person-year that GHC currently
  represents, or put another way, one second for each hour, respectively.}
For another, it means that we are not undertaking significant morphing of GHC
itself:  because this is in large part a student project it must be
kept as simple as reasonably possible in various respects.

\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling must be properly and consistently implemented;
\item The generated code is C.
\end{itemize}

The requirement for tail calling is not primarily for efficiency, rather it
is for a reasonable parallel implementation.  Achieving this, and thereby
sufficiently limiting the use of the native C stack, makes this an interesting
exercise that we discuss at some length.

For feasibility, we restrict ourselves to shared-memory implicit parallelism,
but distributed-memory parallelism should be straightforwardly realizable
using e.g.\ MPI or POSIX sockets.

@@

\subsection{Varying degrees of strictness}
First some words about terminology.  For all degrees of strictness, sharing
(as implied by lazy but not by non-strict) of heap-allocated objects is
preserved so we will not use the term \emph{lazy} and simply refer to
strictness properties.  As is common we will use strict in an operational
sense, meaning that argument(s) are evaluated before being passed in a
function call, rather than the denotational sense, unless stated otherwise.
Similarly, when referring to language semantics we refer by default to the
operational rather than denotational semantics.

GHC 8.0 gives three possibilities: conventional non-strict, constructor
strict, and function- and constructor strict, but there are possible
variations.  The obvious variation is function, but not constructor,
strictness.  In Haskell the distinction between constructors and functions is
somewhat blurred because constructors behave like functions (outside of
pattern matching), but this need not be the case.

\section{Design and Implementation}
Our wish list has much in common with the GHC Haskell implementation, and
indeed GHC is part of our master plan.  In brief, the phases of the GHC
compiler are 1) parsing and type checking; 2) transformation to the GHC
\emph{Core} intermediate representation; 3) analysis and transformation on
Core; 4) transformation of Core to the \emph{STG} intermediate representation;
5) transformation of STG to machine code via one of various imperative
representations (Cmm, LLVM, C).

In short, our plan is to use GHC as our front-end to generate its intermediate
form, STG~\cite{PJ-stockhardware}, then escape to our system from there.
There are various possible ways to accomplish the this: using GHC's option
to dump STG as text, hacking GHC itself and, optionally, using GHC as a
library~\cite{ghc-as-library}.  An alternative would be to start with GHC Core
which we may transition to.

@@

\subsection{Degrees of strictness, take two}
Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG abstract machine.  Our partial description is based on that given
in \emph{EA}.

@@

In the eval-apply model, given application \texttt{f x y}, in the standard
(non-strict) semantics \texttt{f} is first evaluated and the arity of the
underlying \emph{FUN} (user- or system- defined function) or \emph{PAP}
(partial application of a \emph{FUN}) is determined---in the case of
\emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
arguments previously provided.  If the arity is greater than two then a new
\emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
the new \emph{PAP} returned.  If the arity is equal to two then the underlying
\emph{FUN} is tail-called (jumped to) with any previous arguments (in case of
\emph{PAP}) and new arguments \texttt{x} and \texttt{y}.  If the arity is less
than two (it must be at least one, so one in this case) the underlying
\emph{FUN} is called-with-return with any previous arguments (in case of
\emph{PAP}) and new argument \texttt{x}, then the object returned applied to
remaining argument \texttt{y}.

@@

The other dimension we vary is whether or not constructors are strict.
Other dimensions are possible, for example whether \texttt{let} bindings
(heap allocations) are constructed on demand, but we restrict ourselves
to function/application and constructor strictness.

@@

\subsubsection{Proper tail calls.}
Parallelism is achieved by multi-threading; our model is based on Harris et
al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
aspirations~\cite{Marlow:2009}.  To achieve efficient execution they
identified the need for the ability to examine and unwind the stack.  This
requires that use of the C stack be highly constrained if we are to 
avoid complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or
non-portable code to directly manipulate the C stack, complicating an
already non-trivial implementation effort.  We therefore must limit the
use of the C stack to calls to the runtime, in other words, when executing
code generated from the user program (and not in a runtime call), there
must be no growing of (pushing return addresses or data onto) the C stack.

First we establish that this is possible.  By maintaining our own data
and control (continuation) stack all generated C functions can take no
explicit arguments and have return type \emph{void}. If we arrange that
generated C code has only calls (other than runtime calls, which do not call
back to user-program code) in tail call positions, that is, immediately before
a return in the control flow graph, then at non-zero optimization levels both
gcc and Clang/LLVM will generate jumps instead of calls for C calls.  When using
gcc we can also use a fixed set of machine registers for passing/returning
values, much like GHC.

\subsubsection{A taste of the continuation style.}
Now to implementation, which we discuss for a simple case, then the most complex
case.  To clearly distinguish between a notional C call (a syntactic concept)
and a machine-level call (pushing a return address onto the native C stack) we
will refer to the latter as \emph{call-with-return}, and to a machine-level
tail call as \emph{jump}.

First a detail.  Our implementation of the STG abstract machine has an
abstract register called the \emph{current value register} or \emph{CVR}.
Roughly speaking, it can hold the pointer to what is about to be evaluated or
what was just evaluated.  Now consider the evaluation of \texttt{case e of
  <alts>} if call-with-return is disallowed.  If the compiler cannot determine
that evaluating \texttt{e} does not jump somewhere then there must be some
mechanism to get to the \texttt{alts}.  Our solution is to have a separate C
function for the \texttt{alts}, and before executing the code to evaluate
\texttt{e}, to push a continuation---the address of this function and the free
variables of the \texttt{alts}---onto the stack.  Whatever the code for
\texttt{e} might do, ultimately (in the absence of error or non-termination)
the value of \texttt{e} is in the CVR, and some code somewhere reaches its end
and ``returns''---jumps to the address
in the continuation on the top of the stack, thus entering the code for
\texttt{alts}.  Thus continuations direct where to go to, not how to return
to caller as in a stack for an imperative language.

\subsubsection{Is this too much for this paper?}
Similar to what is described in \emph{EA}, in an earlier
implementation we have a generic variadic \emph{stgApply} C function that is
used for applications of non-known functions, and known functions with a
deficit or excess number of the arguments.  The \emph{stgApply} function takes
as arguments the first element of the application (\emph{FUN}, \emph{PAP},
\emph{THUNK}, or \emph{BLACKHOLE}) and its arguments, and if a \emph{THUNK},
evaluates it with a call-with-return to yield a \emph{FUN} or \emph{PAP}.
This allows us to implement a single \emph{stgApply} function as a C function
taking a contiguous block of machine words (unboxed values or pointers) on the
explicit stack.  The implementation of \emph{stgApply} as described in
\emph{EA} performs the same call-with-return, but the scheme described there
uses an auto-generated family of \emph{stgApply} functions parameterized by
number and type (unboxed or pointer) of arguments.  (We used this scheme
initially as well, but because the stack is now fully explicit and the stack
frame is self-describing (e.g., contains a bitmap describing which arguments
are boxed/unboxed for the benefit of the garbage collector), this elaboration
is unnecessary.) Call-with-return was also used in the case that a \emph{FUN}
or \emph{PAP} is provided more than its arity number of arguments, in which
case \emph{FUN} or \emph{PAP} is called-with-return with its arity number
of objects, and the returned object and remaining argument passed (via jump)
to \emph{stgApply}.  Finally, for strict evaluation orders it is straightforward
to evaluate function or constructor arguments one at a time with call-with-return.

>>>>

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{llncs}
%
\usepackage{makeidx}  % allows for indexgeneration
% mkd for preliminary paper notes
\usepackage{enumitem} 
\usepackage{color} % for parenthetical notes while paper is in progress
\usepackage{comment}
%
\begin{document}
%
%mkd \frontmatter          % for the preliminaries
%
\pagestyle{headings}  % switches on printing of running heads
%
%mkd \mainmatter              % start of the contributions
%
\title{Automatic Parallelization and Transparent Fault Tolerance\\
       (New project paper)}
%
\titlerunning{Automatic Parallelization}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Kei Davis\inst{1} \and
        Dean Prichard\inst{1} \and
        David Ringo\inst{1,2} \and
        Loren Anderson\inst{1,3} \and
        Jacob Marks\inst{1,4}\thanks{All project contributors, past and present, are listed.}
}
%
\authorrunning{Kei Davis et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
%\tocauthor{Ivar Ekeland, Roger Temam, Jeffrey Dean, David Grove,
%Craig Chambers, Kim B. Bruce, and Elisa Bertino}
%
\institute{Los Alamos National Laboratory, Los Alamos, NM, USA\\
\email{kei.davis@lanl.gov},\\ WWW home page:
\texttt{http://ccsweb.lanl.gov/\homedir kei/}
\and
University of New Mexico,
Albuquerque, NM, USA
\and
North Dakota State University,
Fargo, North Dakota, USA
\and
New Mexico Institute of Mining and Technology,
Socorro, New Mexico, USA
}

@@

\begin{abstract}
The claimed trend is an increasing appreciation and practice of functional
programming in scientific computing, and in particular strict (by default)
functional programming, in various guises.  Our project is the demonstration
of a light-weight, higher-order, polymorphic, pure functional language
implementation in which we can experiment with automatic parallelization
strategies and varying degrees of default strictness.  Non-strictness
constrains the degree of potential parallelism, that is, we are not exploring
speculative evaluation.  A secondary goal is to experiment with mechanisms
for transparent fault tolerance, another trend in scientific computing, and
by transitivity functional programming.
\\
\emph{Draft paper notes:}
\begin{enumerate}[before=\itshape]
  \item We are reporting work in progress.  At the time of writing we have a
    complete serial implementation, a primitive proof-of-concept parallel
    implementation, and are currently developing a more realistic parallel
    runtime.  We expect to have preliminary results from our new parallel
    implementation by the time of the symposium.  The fault-tolerance mechanisms
    remain future work.

@@

  \item Hopefully in the spirit of a student-friendly symposium, for this
    draft version we have chosen a more informal, conversational style of
    presentation than necessarily appropriate for a final paper, have
    highlighted student involvement and its influence on our approach, and
    included perhaps gratuitous detail about student contributions in particular.
  \end{enumerate}

\keywords{strict pure functional programming, automatic parallelization, Haskell, STG,
transparent fault tolerance.}

@@

\subsection{The trend: strict but still pure functional programming in HPC}
Interest in functional programming for HPC in general is not new, as evidenced
by the Workshop on Functional High Performance Computing, held in conjuction
with the International Conference on Functional Programming, being now in its fifth
consecutive year.  A more precise characterization of the trend is that there
is a small but growing understanding by practicing computational scientists
that \emph{pure functional semantics}, in some form, is essential to reining
in the complexity of these ever-evolving scientific codes and informing the
designs of new ones.  Such semantics are expressed in a number of ways as
highlighted following.

@@

\subsubsection{Use a pure functional language.}
Some twenty years after the US Department of Energy laboratory complex was
presented with the most highly parallel (and thereby performant)
implementation of a pure functional language to date~\cite{Davis96}, and
collectively had no conceptual understanding of what it was, much less why it
might be useful or important, there is now a small number of laboratory HPC
practioners who are developing an appreciation for pure functional programming
in Haskell.  Still, they see their needs as different than other Haskell
users: performance is paramount, unboxed arrays are a (perhaps \emph{the})
primary data structure, and non-strictness or laziness is more of a
nuisance--even a serious hindrance--than a convenience, much less ever
algorithmically essential in the sense of Bird's \emph{repmin}~\cite{Bird84},
Johnsson's general attribute grammars~\cite{Johnsson87}, or other instances
of \emph{tying the knot}~\cite{tying-the-knot}.

A common complaint among these aspiring high-performance functional
programmers is the need to get strictness just right: to remove space leaks,
achieve acceptable performance, get communication (e.g.\ MPI) and IO to work
as they expect, and so on.  More specifically, they expect to be able to
reason about space and time usage, and order of execution, using their
existing mental models. While they are happy to write code that is strongly
reminiscent of the numerical examples in Hughes' classic piece (whether
they've read it or not)~\cite{Hughes89}, they would also be happy to write in
a slightly modified style that does not require non-strictness---after all,
that's how they've been thinking their entire careers.

Much anecdotal evidence suggests that many others have similar difficulties
with semantics that are lazy (or non-strict) by default, with the most telling
recent concrete evidence being the push for, and realization of, the
\texttt{Strict} and \texttt{StrictData} language extensions newly available in
GHC 8.0.  Quoting the documentation:
\begin{quote}
  High-performance Haskell code (e.g.\ numeric code) can sometimes be littered
  with bang patterns, making it harder to read. The reason is that laziness
  isn't the right default in this particular code, but the programmer has no
  way to say that except by repeatedly adding bang
  patterns~\cite{strict-strictdata}.
\end{quote}
It is a fact that \emph{HPC practioners are by and large not computer
  scientists}, they are domain scientists (physicists, chemists, etc.) or
computational scientists specialized in numerical or numerically-oriented
algorithmic methods (e.g., Lagrangian, Eulerian, adaptive mesh refinement).
The concept of a space leak, for example, caused by other than a missing
explicit deallocation statement (C \emph{free}, etc.) is difficult to
understand and not something they particularly want to try to understand.
%Hence, we eschew explicit strictness annotations for varying uniform degrees
%of strictness.

More generally, \emph{task-based} parallelism is becoming increasingly popular
for HPC, often in conjunction with a higher-level (typically inter-node) bulk
synchronous model typically implemented with the Message Passing Interface
(MPI)~\cite{MPI}.  In the imperative world, example systems include the
aforementioned Legion, Intel's Threading Building Blocks~\cite{Reinders:2007},
and the HPX C++11/14 extension~\cite{HPX}.  While the concept of task is
somewhat vague, the general idea is a unit of computation with no or
minimal/constrained side effects.  A function in a pure functional language
could be regarded as the essence of the concept of task.  All of these
systems are of course in languages with strict function-call semantics.

In the quest for computing platforms approaching exascale, it is widely
recognized that new mechanisms for fault tolerance will need to be built into
the software stack.  Current practice in HPC is checkpoint/restart, wherein at
various points in time sufficient program state is dumped to backing store such
that the program, in case of a fault, can be (manually) restarted from the
latest stored state.  Most implementations are explicit, though there has been
some success with transparent (invisible to the programmer) systems.  Wiring
checkpoint/restart into a large-scale application is not only burdensome and
error-prone, it can severely warp the engineering of a large-scale application
because it must be designed such that essential state can be readily captured.
And now, it is widely believed that the scalability of checkpoint/restart may
be reaching its end.

In the imperative world side-effect-free tasks could be safely restarted after
failure, needing only their well-defined inputs and not arbitrary access to
mutable global state.  The Legion system is being augmented with this
capability, and in such a way that it will be largely transparent to the
programmer.  Ericsson's quasi-functional language Erlang and its runtime
system were designed for fault tolerance~\cite{Cesarini:2009}.  It has long
been recognized that the pure parts of a functional program (excluding, e.g.,
the I/O monad in Haskell) could be safely restarted.  The recent work of
Stewart impressively demonstrates this on a large-scale distributed-memory
system~\cite{Stewart:2013}.

We claim, then, that there is a trend, in HPC at least, towards
\emph{strict-by-default pure functional programming}, whether as imposed by
programmer discipline, a parallel runtime system, or the language itself
(e.g.\ strict Haskell).  Second, we claim that there is a trend in
high-performance computing, and by transitivity (in addition to Stewart's
work) in functional programming, towards transparent fault tolerance.

@@

\section{Project and Goals}
Our project is the light-weight implementation of a pure, higher-order,
polymorphic, functional language and runtime system with which we can
experiment with automatic parallelization strategies with varying degrees of
language strictness, and secondarily, with mechanisms for transparent fault
tolerance.  Light-weight has several implications.  First and foremost it
means that \emph{we are not attempting to compete with GHC in any
  respect}---succumbing to the temptation of recreating GHC language, type
system, and runtime features, or significant code optimization, would simply
divert us from our central goals.\footnote{We estimate that we will have, to
  within an order of magnitude, one person-hour of total available development
  time for each person-year that GHC currently represents, or put another way,
  one second for each hour, respectively.}  For another, it means that we are
not undertaking significant morphing of GHC itself: because this is in large
part a student project it must be kept as simple as reasonably possible in
various respects.  Our initial requirements are that
\begin{itemize}
\item The implementation must be feasible in terms of effort;
\item The language must be strongly typed, implicitly or explicitly.  More
  specifically, the language should explicitly support both Hindley-Milner
  polymorphism and unboxed types~\cite{Jones:1991};
\item Straightforward direct linkage with C, preferably without a sophisticated
  foreign function interface, is essential;
  \item The language should be \emph{currying-friendly} as described by Marlow
and Peyton Jones~\cite{Marlow:2004};
\item Choice of degree of strictness should be easily selected;
\item Sharing, in the sense of what \emph{lazy} means vs.\ merely non-strict,
  should be preserved for all degrees of strictness;
\item Tail calling must be properly and consistently implemented;
\item The generated code is C.
\end{itemize}

For feasibility, we restrict ourselves to shared-memory implicit parallelism.
The requirement for tail calling is not primarily for efficiency, rather it
is for a reasonable parallel implementation.  Achieving this, and thereby
sufficiently limiting the use of the native C stack, makes this an interesting
exercise that we discuss at some length.

@@

\section{Design and Implementation}
Our wish list has much in common with the GHC Haskell implementation, and
indeed GHC is part of our master plan.  In brief, the phases of the GHC
compiler are 1) parsing and type checking; 2) transformation to the GHC
\emph{Core} intermediate representation~\cite{Sulzmann:2007,ghc-core}; 3)
analysis and transformation on Core; 4) transformation of Core to the
\emph{STG} intermediate representation~\cite{PJ-stockhardware}; 5)
transformation of STG to machine code via one of various imperative
representations (Cmm, LLVM, C).

In short, our plan is to use GHC as our front-end to generate one of its
intermediate forms, STG or Core, then escape to our system from there.  There
are various possible ways to accomplish the this: using GHC's option to dump
STG or Core as text, hacking GHC itself and, optionally, using GHC as a
library~\cite{ghc-as-library}.  For now we start with our own STG, with
bridging the gap with GHC scheduled to commence soon.

@@

\subsection{Degrees of strictness}
First some words about terminology.  For all degrees of strictness, sharing
(as implied by lazy but not by non-strict) of heap-allocated objects is
preserved so we will not use the term \emph{lazy} and simply refer to
strictness properties.  As is common we will use strict in an operational
sense, meaning that argument(s) are evaluated before being passed in a
function call, rather than the denotational sense, unless stated otherwise.
Similarly, when referring to language semantics we refer by default to the
operational rather than denotational semantics.

GHC 8.0 gives three possibilities: conventional non-strict, constructor
strict, and function- and constructor strict, but there are possible
variations.  The obvious variation is function, but not constructor,
strictness.  In Haskell the distinction between constructors and functions is
somewhat blurred because constructors behave like functions (outside of
pattern matching), but in STG the distinction is clear.

Other possible variations concern the notional arity of the underlying
function being applied.  This is readily explained in terms of the operation
of the STG abstract machine.  Our partial description is based on that given
in \emph{EA}.

@@

In the eval-apply model, given application \texttt{f x y}, in the standard
(non-strict) semantics \texttt{f} is first evaluated and the arity of the
underlying \emph{FUN} (user- or system- defined function) or \emph{PAP}
(partial application of a \emph{FUN}) is determined---in the case of
\emph{PAP} the arity is that of the underlying \emph{FUN} less the number of
arguments previously provided.  If the arity is greater than two a new
\emph{PAP} is created with arguments \texttt{x} and \texttt{y} inserted and
the new \emph{PAP} returned.  If the arity is equal to two the underlying
\emph{FUN} is tail-called (jumped to) with any previous arguments (in case of
\emph{PAP}) and new arguments \texttt{x} and \texttt{y}.  If the arity is less
than two (it must be at least one, so one in this case) the underlying
\emph{FUN} is called-with-return with any previous arguments (in case of
\emph{PAP}) and new argument \texttt{x}, then the object returned applied to
remaining argument \texttt{y}.

@@

The other dimension we vary is whether or not constructors are strict.  Other
dimensions are possible, for example whether \texttt{let} bindings are strict,
but we restrict ourselves to function/application and constructor strictness.

@@

\subsubsection{Proper tail calls.}
Parallelism is achieved by multi-threading; our model is based on Harris et
al.'s.~\cite{Harris:2005}; later developments by Marlow et al.\ are beyond our
aspirations~\cite{Marlow:2009}.  To achieve efficient parallel execution they
identified the need for the ability to examine and unwind the stack.  This
requires that use of the C stack be highly constrained if we are to 
avoid complex and ugly hacks using, e.g., \emph{setjmp/longjmp}, or
non-portable code to directly manipulate the C stack, complicating an
already non-trivial implementation effort.  We therefore must limit the
use of the C stack to calls to the runtime, in other words, when executing
code generated from the user program (and not in a runtime call), there
must be no growing of (pushing return addresses or data onto) the C stack.

First we establish that this is possible in principle.  By maintaining our own
data and control (continuation) stack all generated C functions can take no
explicit arguments and have return type \emph{void}. If we arrange that
generated C code has only calls (other than runtime calls, which do not call
back to user-program code) in tail call positions, that is, immediately before
a return in the control flow graph, then at non-zero optimization levels both
gcc and Clang/LLVM will generate jumps instead of calls for C calls.  When
using gcc we can also use a fixed set of machine registers for
passing/returning values, much like GHC.

\subsubsection{A taste of the continuation style.}
Now to implementation.  To clearly distinguish between a notional C call (a
syntactic concept) and a machine-level call (pushing a return address onto the
native C stack) we will refer to the latter as \emph{call-with-return}, and to
a machine-level tail call as \emph{jump}.

First some details.  Our implementation of the STG abstract machine has an
abstract register called the \emph{current value} register or
\emph{CV}.\footnote{There are actually two registers, one for pointers and
  one for unboxed values, but for simplicity we minimize discussion about
  unboxed values.}  Roughly speaking, it can hold the pointer to what is about
to be evaluated or what was just evaluated.  For example, when initiating
evaluation of a \emph{THUNK} object, \emph{CV} contains a pointer to that
object; it is much like the \emph{self} pointer in C++.  On completion of
evaluation \emph{CV} will contain a pointer to the object to which the
\emph{THUNK} evaluated.  A \emph{THUNK} will not return another
\emph{THUNK}---an evaluation is always ``big step'' to (at least) WHNF, that
is, to a \emph{FUN}, \emph{PAP}, or \emph{CON}.
% These latter, if evaluated, simply return through the continuation stack,
%effectively returning \emph{self}.

All continuations contain a code pointer, optionally some data, and layout
information.  These continuations are maintained on a stack.  When the
evaluation of an object is completed its return is by jumping through the code
pointer of the continuation on the top of the stack (``returning through the
continuation stack'').  For example, the argument frame for a function is one
type of continuation and its associated code simply pops the continuation off
the stack and returns through the next continuation on the stack.

Before evaluating a user program an initial continuation is pushed onto
the stack that points to a C function that performs whatever post-user-program
processing might be wanted, such as displaying the result of the program in
\emph{CV}\@.  The driver then jumps to the entry point of the user program,
which will ultimately return through this continuation.  This makes clear
that continuations direct \emph{where to go to}, not \emph{how to return
to the caller} as in a stack for an imperative language, despite the use of
the term ``return.''

Now consider the evaluation of \texttt{case e of x \{ alts \}} if call-with-return
is disallowed.  If the compiler cannot determine that evaluating \texttt{e}
does not jump somewhere then there must be some mechanism to get to the
\texttt{alts}.  Our solution follows \emph{EA}, using a separate C function
for the \texttt{alts}, and before executing the code to evaluate \texttt{e},
to push a continuation---the address of this function and the free variables
of the \texttt{alts}---onto the stack.  Whatever the code for \texttt{e} might
do, ultimately (in the absence of error or non-termination) the value of
\texttt{e} is in \emph{CV}, and some code somewhere reaches its end and returns
through the continuation stack, thus entering the code for \texttt{alts},
which immediately binds \emph{x} to the value in \emph{CV}.

\emph{EA} describes a family of generic C-{}- \emph{stgApply} runtime
functions used for applications of non-known (not statically identifiable)
functions, and applications of known functions to other than their arity
number of arguments.  This family is parameterized by the number of actual
arguments in the application and whether each is boxed or unboxed.  Because
we have an explicit stack we are able to use a single C \emph{stgApply} function
taking a contiguous block of machine words (unboxed values or
pointers) on the explicit stack.  Some modifications are required, however,
because \emph{EA} uses call-with-return.  Specifically, when a \emph{FUN} or is
\emph{PAP} applied to more than its arity number of arguments
a call-with-return to the (underlying) \emph{FUN} is made with the arity
number of arguments, then the appropriate \emph{stgApply} function is jumped
to, passing the newly returned object and the remaining arguments.  Our
solution to this entails passing \emph{stgApply} the object to be applied in
\emph{CV}, together with an extra argument indicating which of the object
to be applied and its boxed arguments have been already been evaluated (by
\emph{stgApply}).  If the object to be applied, or some argument, remains to
be evaluated (because of the strictness level), it is jumped to, with
return as ever through the continuation stack.  This is actually fairly
efficient because we can arrange that the stack frame for \emph{stgApply}
is reused, so except for the indirect jumps on return this is essentially
a loop.  For the cases of arguments in excess of the arity of the object
being applied, in effect \emph{stgApply}'s argument frame is popped and
two continuations are pushed, first one to jump to \emph{stgApply} with
the excess arguments, and the next an argument frame for the object to
applied, then the code for the object to be applied is jumped to.
